---
title: "Bayesian Ideal Point Estimation of the Italian Chamber of Deputies"
author: "Lodovico Monteforte"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(pscl)
library(ggplot2)
library(coda)
library(tidyr)
library(dplyr)
library(patchwork)
library(knitr)
library(scales)
```

---

## Introduction

This project estimates the ideological positions of members of the Italian
Chamber of Deputies using the Bayesian ideal point model introduced by Clinton,
Jackman and Rivers (2004).^[Clinton, J., Jackman, S., & Rivers, D. (2004).
The Statistical Analysis of Roll Call Data. *American Political Science Review*,
98(2), 355–370. https://doi.org/10.1017/S0003055404001194] The method recovers
latent ideological positions — ideal points — from the observed pattern of
Yea/Nay votes on roll calls, treating the voting record as data and ideology as
an unobserved quantity to be estimated.

The Italian institutional setting offers a particularly rich testing ground.
The Chamber has 400 members (compared to 100 in the US Senate), supports a
highly fragmented multi-party system, and has experienced substantial
governmental instability across legislatures. In practice this means that the relative distance
between any two candidates changes as new government coalitions come into being.  
Two legislatures are analysed:

- XIX Legislature (13 October 2022 — present): the current centre-right
  government led by Giorgia Meloni, with a clear and stable majority coalition.
  This serves as a baseline case where a strong ideological separation is
  expected a priori.

- XVIII Legislature (23 March 2018 — 12 October 2022): a more complex case
  involving three distinct governments (the M5S-Lega coalition, the
  M5S-PD-IV coalition, and the Draghi technocratic government).

The analysis is carried out using the `pscl` package^[https://rdocumentation.org/packages/pscl/versions/1.5.9]
by Simon Jackman — one of the original authors of the paper — which implements
a Gibbs sampler. The following sections present the
theoretical foundations of the model, the data pipeline, and the results for
both legislatures.

---

## Theory

### The Spatial Voting Model

A classical building block of positive political theory is the **spatial voting
model**. Each legislator is modelled as a rational utility maximiser with an
ideal point in a $d$-dimensional policy space: a point representing their
most preferred policy position. The closer a proposed policy is to a
legislator's ideal point, the higher their utility from supporting it.

We assume $n$ legislators vote on $m$ roll calls. For each roll call $j$,
legislator $i$ chooses between a Yea position ($\zeta_j$) and a Nay position
($\psi_j$) in the policy space. The utility functions are:

$$U_i(\zeta_j) = -\|x_i - \zeta_j\|^2 + \eta_{ij}$$
$$U_i(\psi_j) = -\|x_i - \psi_j\|^2 + \nu_{ij}$$

where $x_i \in \mathbb{R}^d$ is the ideal point of legislator $i$, and
$\eta_{ij}$, $\nu_{ij}$ are stochastic error terms capturing unobserved
utility. A legislator votes Yea ($y_{ij} = 1$) if:

$$P(y_{ij} = 1) = P(U_i(\zeta_j) > U_i(\psi_j))$$


Expanding the Euclidean distance terms
($\|a - b\|^2 = a'a - 2a'b + b'b$):

$$\|x_i - \psi_j\|^2 - \|x_i - \zeta_j\|^2 = 2(\zeta_j - \psi_j)'x_i +
\psi_j'\psi_j - \zeta_j'\zeta_j$$

Assuming the error difference $\epsilon_{ij} = \nu_{ij} - \eta_{ij}$ is
distributed $N(0, \sigma_j^2)$ and normalising by $\sigma_j$:

$$P(y_{ij} = 1) = \Phi \left(
\frac{2(\zeta_j - \psi_j)'}{\sigma_j}x_i -
\frac{\zeta_j'\zeta_j - \psi_j'\psi_j}{\sigma_j}
\right)$$

We now define:

- $\beta_j = \frac{2(\zeta_j - \psi_j)}{\sigma_j}$: as the discrimination
  parameter, capturing how strongly vote $j$ separates legislators along
  the ideological dimension. A large $|\beta_j|$ indicates a highly
  partisan vote; a near-zero $\beta_j$ indicates a vote on which ideology
  plays little role.

- $\alpha_j = \frac{\zeta_j'\zeta_j - \psi_j'\psi_j}{\sigma_j}$: the
  difficulty parameter, capturing the overall propensity to vote Yea
  regardless of ideology.

The probability of a Yea vote becomes a standard probit model:

$$P(y_{ij} = 1) = \Phi(x_i'\beta_j - \alpha_j)$$

With the joint likelihood for the observed vote matrix $Y$, assuming independence
across legislators and roll calls, is:

$$L(\mathbf{B}, \alpha, \mathbf{X} \mid Y) = \prod_{i=1}^{n} \prod_{j=1}^{m}
\Phi(x_i'\beta_j - \alpha_j)^{y_{ij}}
\left[1 - \Phi(x_i'\beta_j - \alpha_j)\right]^{1 - y_{ij}}$$

where $\mathbf{X}$ is the $n \times d$ matrix of ideal points and $\mathbf{B}$
is the $m \times d$ matrix of discrimination parameters.

Clinton et al. (2004) notice that the posterior
$p(\mathbf{X}, \mathbf{B}, \alpha \mid Y)$ has a tractable conditional
structure under conjugate priors, enabling estimation via Gibbs sampling.

The Gibbs sampler iteratively draws from the conditional posteriors:

1. For $t = 1$ to $T$:
2. Sample $x_i^{(t+1)}$ from $p(x_i \mid \mathbf{B}^{(t)}, \alpha^{(t)}, Y)$
   for all $i$
3. Sample $(\beta_j^{(t+1)}, \alpha_j^{(t+1)})$ from
   $p(\beta_j, \alpha_j \mid \mathbf{X}^{(t+1)}, Y)$ for all $j$
4. Update $\theta^{(t+1)} \leftarrow (x^{(t+1)}, \beta^{(t+1)}, \alpha^{(t+1)})$

What makes it tractable here is the Albert-Chib data
augmentation: by introducing latent continuous variables $z_{ij}$ such that
$y_{ij} = \mathbf{1}[z_{ij} > 0]$ with $z_{ij} \sim N(x_i'\beta_j - \alpha_j, 1)$,
the conditional posteriors for both ideal points and bill parameters become
Gaussian, enabling closed-form sampling at each step.

### Identification

The model has two sources of non-identification:

1. **Sign flip**: multiplying all $x_i$ by $-1$ and all $\beta_j$ by $-1$
   leaves the likelihood unchanged.
2. **Scale**: multiplying all ideal points by a constant $c$ and dividing
   all $\beta_j$ by $c$ leaves the likelihood unchanged.

Both are resolved by fixing the ideal points of two legislators to known
values. We use **Fratoianni Nicola** (left, fixed at $-1$) and
**Lollobrigida Francesco** (right, fixed at $+1$) as anchors. Both are
long-serving members of their respective parties with stable and well-known
ideological positions, making them suitable anchor choices.
---

## Data

### Source and Scraping Pipeline

Vote data was obtained from the Italian Chamber of Deputies open data portal
(`dati.camera.it`).
For each legislature, we queried all final votes (`votazioneFinale = 1`),
excluding procedural votes. For each vote, we retrieved the legislator's name,
vote expression (Yea/Nay/Abstain), and party group affiliation at the date of
the vote — the latter using date filters to correctly assign party membership
for legislators who switched groups mid-legislature.

The scraping pipeline is implemented in Python and supports automatic resume
on interruption. The full scrape takes approximately 20 minutes per
legislature.

### Preprocessing

The raw dataset (one row per legislator per vote) was transformed into a
legislator × vote binary matrix where 1 = Yea, 0 = Nay, and NA = absent
or abstaining. Party names were standardised from the verbose raw labels to
short identifiers (e.g. `"FRATELLI D'ITALIA (FDI) (18.10.2022"` → `"FdI"`).

Legislators who switched party mid-legislature were collapsed onto a single
identifier, retaining their full voting record but assigned to their last
party of affiliation. In theory this could create outliers for extreme
party-switchers, but such cases should be rare and do not affect model mechanics
since ideal points are estimated from votes, not party labels.

### Sample Reduction

Before estimation, the `pscl::ideal` function applies a lop filter
(`lop = 0.05`): roll calls where fewer than 5% of legislators voted on
the minority side are dropped as uninformative. The table below reports
the approximate effect of this filter on both datasets. 

```{r sample-reduction}
red_18 <- read.csv("sample_reduction_18.csv")
red_19 <- read.csv("sample_reduction_19.csv")
reduction_df <- rbind(red_19, red_18)

knitr::kable(reduction_df,
             col.names = c("Legislature", "Legislators (raw)",
                           "Legislators (no votes)",
                           "Votes (raw)", "Votes dropped (lop < 0.05)",
                           "Votes kept"),
             caption = paste("Sample reduction after lop = 0.05 filter.",
                             "Note: counts are approximate — exact filtering",
                             "is performed by ideal() in R."))
```

---

## Results: XIX Legislature (2022–present)

```{r load-19, cache=TRUE}
load("session_19.RData")
load("sensitivity_19.RData")

all_parties <- sort(unique(c(
  "FdI", "FI", "IPF-IC", "M5S", "MISTO", "IV", "Lega", "PD", "LEU-ART1-SI",
  "PD-IDP", "AZ-PER-RE", "AVS", "IV-C-RE", "NM"
)))

master_colors <- setNames(
  scales::viridis_pal(option = "turbo")(length(all_parties)),
  all_parties
)
```

### Ideological Distribution

```{r plot-a-19, fig.cap="Posterior mean ideal points for all XIX legislature members. Vertical jitter is purely cosmetic to avoid overplotting. The vertical line marks the ideological centre (0)."}
plot_df_19 <- data.frame(
  legis.name = rownames(id_19$xbar),
  Estimate   = as.numeric(id_19$xbar[, 1]),
  stringsAsFactors = FALSE
)
plot_df_19 <- merge(plot_df_19, data19[, c("legis.name", "partito")],
                    by = "legis.name", all.x = TRUE)

set.seed(42)
ggplot(plot_df_19, aes(x = Estimate, y = 0, color = partito)) +
  geom_jitter(size = 2, alpha = 0.65, height = 0.15) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50") +
  geom_vline(xintercept = 0, linewidth = 0.8, color = "black") +
  scale_color_manual(values = master_colors) +
  theme_minimal() +
  labs(title    = "Ideological Distribution — XIX Legislature",
       subtitle = "Posterior means | Anchors: Fratoianni = -1, Lollobrigida = +1",
       x        = "← Left          Right →",
       y        = "", color = "Party") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(), legend.position = "bottom")
```

The model recovers a clean left-right separation consistent with the known
political landscape of the XIX legislature. The right-wing governing coalition
(FdI, Lega, FI, NM) clusters compactly on the right, while the opposition
(PD-IDP, AVS, M5S) occupies the left. AVS is correctly placed further left
than PD-IDP. The M5S occupies a hard-left position.

### Bill Parameters

Another parameter of interest is the discrimination parameter $\beta_j$. This measures
how strongly each vote separates left from right legislators. Its distribution tells us
how well voting patterns map onto a single ideological axis, and whether a single
ideological dimension is sufficient to explain voting behaviour. A concentration of
near-zero $\beta_j$ values would indicate that many votes carry little ideological
signal; systematic departures from a unimodal legislator distribution would instead
suggest the need for an additional dimension beyond left and right.



```{r plot-d-19, fig.width=12, fig.height=5, fig.cap="Left: distribution of bill discrimination parameters. Right: discrimination vs difficulty for all votes."}
beta_df_19 <- data.frame(
  vote_id = 1:nrow(id_19$betabar),
  beta    = id_19$betabar[, 1],
  alpha   = id_19$betabar[, 2]
)
noisy_pct_19 <- round(mean(abs(beta_df_19$beta) < 0.5) * 100, 1)

p_hist <- ggplot(beta_df_19, aes(x = beta)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "steelblue", color = "white", alpha = 0.6) +
  geom_density(color = "black", linewidth = 0.8, adjust = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  theme_minimal() +
  labs(title    = "Distribution of β (XIX)",
       subtitle = paste0("Votes with |β| < 0.5: ", noisy_pct_19, "%"),
       x = "β (Discrimination)", y = "Density")

hull_idx_19 <- chull(beta_df_19$beta, beta_df_19$alpha)
hull_df_19  <- beta_df_19[hull_idx_19, ]

p_scatter <- ggplot(beta_df_19, aes(x = beta, y = alpha)) +
  geom_polygon(data = hull_df_19, fill = "steelblue", alpha = 0.15,
               color = "steelblue", linewidth = 0.5) +
  geom_point(alpha = 0.3, size = 1.2, color = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.6) +
  theme_minimal() +
  labs(title    = "β vs α (XIX)",
       subtitle = "Each point is one roll-call vote",
       x = "β (Discrimination)", y = "α (Difficulty)")

p_hist + p_scatter
```

In our data, the trimodal structure of the $\beta$ distribution reflects three
behavioural types of votes: highly partisan votes driven by the right-wing majority
(large positive $\beta$), opposition-sponsored votes where the left consistently votes
Yea and the right Nay (large negative $\beta$), and procedural or consensus votes where
ideology plays little role (near-zero $\beta$). Only `r noisy_pct_19`% of votes have
$|\beta| < 0.5$, indicating that the 1D specification captures the vast majority of the
ideological signal in the data.

Figure X plots $\beta$ against $\alpha$. A large share of votes falls in the first
quadrant, combining high discrimination with a Yea being the the majority-favored outcome. 
This is coherent with the relative strength and cohesion of the current government.


### A side note on convergence and party mechanics

The model was run for 150,000 iterations. Assessing convergence via the Geweke
diagnostic — which tests whether the mean of the first 10% of the chain equals
the mean of the last 50% — reveals that a large share of individual ideal point
parameters do not converge (see table below). This is a known limitation of the
model: we have to assume non strategic, truthful voting. 
In a setting like the Italian parliament operates under strong party whipping, producing
low heterogeneity in voting behaviour within parties and making precise
estimation of *individual* ideal points difficult regardless of chain length.
Nevertheless, the estimates remain informative about the relative positioning of
parties along the left-right spectrum. Since legislators within the same party
vote near-identically, what the model recovers is a reliable signal of
*inter-party* ideological distances, which is the quantity of primary interest
in our analysis.

```{r geweke-19, cache=TRUE}
leg_19   <- as.mcmc(id_19$x[, , 1])
beta_19  <- as.mcmc(id_19$beta[, , 1])
alpha_19 <- as.mcmc(id_19$beta[, , 2])

gw_leg_19   <- geweke.diag(leg_19)
gw_beta_19  <- geweke.diag(beta_19)
gw_alpha_19 <- geweke.diag(alpha_19)

geweke_df_19 <- data.frame(
  Parameter  = c("Ideal Points", "β (Discrimination)", "α (Difficulty)"),
  Total      = c(id_19$n, id_19$m, id_19$m),
  Failed     = c(sum(abs(gw_leg_19$z)   > 1.96),
                 sum(abs(gw_beta_19$z)  > 1.96),
                 sum(abs(gw_alpha_19$z) > 1.96)),
  Pass_Rate  = paste0(round((1 - c(
                 mean(abs(gw_leg_19$z)   > 1.96),
                 mean(abs(gw_beta_19$z)  > 1.96),
                 mean(abs(gw_alpha_19$z) > 1.96))) * 100, 1), "%")
)

knitr::kable(geweke_df_19,
             col.names = c("Parameter", "Total", "Failed (|Z| > 1.96)",
                           "Pass Rate"),
             caption = "Geweke Convergence Diagnostics — XIX Legislature (1D)")
```

An important caveat applies to legislators with very few recorded votes. A
striking case is Prime Minister Giorgia Meloni, who cast no votes in this
legislature — a common pattern among Prime Ministers. With no data, the
posterior is entirely determined by the prior $\mathcal{N}(0, 1)$, and the
chain will appear to converge trivially since it never moves away from the
prior. Her Geweke Z-score is
`r round(gw_leg_19$z[grep("MELONI GIORGIA", names(gw_leg_19$z))], 3)`,
which is
`r ifelse(abs(gw_leg_19$z[grep("MELONI GIORGIA", names(gw_leg_19$z))]) < 1.96, "below", "above")`
the 1.96 threshold — a formally convergent but entirely uninformative result.

```{r trace-19, fig.width=12, fig.height=5, fig.cap="Trace plots for Meloni (no data, prior-dominated)."}
idx_meloni   <- grep("MELONI GIORGIA",   colnames(leg_19))

par(mfrow = c(1, 2))
coda::traceplot(leg_19[, idx_meloni],
                main = "Trace: Meloni Giorgia (0 votes)",
                col = "#377EB8", ylab = "Ideal Point")
```

### Sensitivity to Bill Prior Specification

Following Jackman (2001),^[Jackman, S. (2001). Multidimensional Analysis of
Roll Call Data via Bayesian Simulation. *Political Analysis*, 9(3), 227–241.
https://doi.org/10.1093/polana/9.3.227] we assess sensitivity to the bill
prior variance $\sigma^2_\beta$ by running three short chains under different
specifications: Tight ($\sigma^2_\beta = 1$), Default ($\sigma^2_\beta = 25$),
and Flat ($\sigma^2_\beta = 625$). If the data are sufficiently informative,
ideal point estimates should be stable across specifications.
As can be seen below, tighter priors on $\beta_j$ compress the discrimination
parameters toward zero, which in turn shrinks the estimated spread of ideal
points symmetrically — pulling both left-leaning and right-leaning parties
toward the center. Crucially, however, this compression is approximately
uniform across the political spectrum: the rank ordering of parties and their
relative distances are preserved across all three specifications. We therefore
conclude that the choice of $\sigma^2_\beta$ does not materially affect the
substantive conclusions, at least with respect to the inter-party positional
comparisons that are the primary object of our analysis.

```{r plot-c-19, fig.width=12, fig.height=5, fig.cap="Each point is one legislator. Points on the dashed identity line are unaffected by prior choice."}
ideal_comp_19 <- data.frame(
  legis.name = rownames(id_1_19$xbar),
  Tight      = as.numeric(id_1_19$xbar[, 1]),
  Default    = as.numeric(id_2_19$xbar[, 1]),
  Flat       = as.numeric(id_3_19$xbar[, 1]),
  stringsAsFactors = FALSE
)
ideal_comp_19 <- merge(ideal_comp_19, data19[, c("legis.name", "partito")],
                       by = "legis.name", all.x = TRUE)

p_c1_19 <- ggplot(ideal_comp_19, aes(x = Tight, y = Default, color = partito)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = FALSE, color = "red",
              linewidth = 0.8, aes(group = 1)) +
  scale_color_manual(values = master_colors, guide = "none") +
  theme_minimal() +
  labs(title = "Tight (Var=1) vs Default (Var=25)",
       x = "Ideal Point (Tight)", y = "Ideal Point (Default)")

p_c2_19 <- ggplot(ideal_comp_19, aes(x = Default, y = Flat, color = partito)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = FALSE, color = "red",
              linewidth = 0.8, aes(group = 1)) +
  scale_color_manual(values = master_colors) +
  theme_minimal() +
  labs(title = "Default (Var=25) vs Flat (Var=625)",
       x = "Ideal Point (Default)", y = "Ideal Point (Flat)")

p_c1_19 + p_c2_19 +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  plot_annotation(
    title    = "Sensitivity to Bill Prior Specification — XIX Legislature",
    subtitle = "Dashed line: identity | Red line: OLS fit")
```

### Data, Uncertainty and Precision


Another point of interest is convergence under low participation. To examine how
well the model characterises legislators with few recorded votes, we identify the
three least active members for each party and plot their 95% credible intervals.

Legislators with few votes have posteriors dominated by the prior, yielding wide
and potentially uninformative estimates. What we observe below, however, is a
clear separation of the main parliamentary blocs even in this low-information
setting. Coherently with the current political environment, the model
distinguishes a solid and cohesive right-wing bloc from a more heterogeneous
left, reflecting the asymmetry in party discipline discussed above. Two members
of the right-wing bloc do display very wide credible intervals positioned on the
left side of the scale, suggesting their posteriors are almost entirely
prior-dominated a consequence of near-complete absence from the chamber.

```{r plot-b-19, fig.height=8, fig.cap="Grey circle size = 95% CI half-width. Colored dot = posterior mean. The vertical line marks the ideological centre."}
ci_df_19 <- data.frame(
  legis.name = rownames(id_19$xbar),
  mean       = as.numeric(id_19$xbar[, 1]),
  lower      = apply(id_19$x[, , 1], 2, quantile, 0.025),
  upper      = apply(id_19$x[, , 1], 2, quantile, 0.975),
  stringsAsFactors = FALSE
)
ci_df_19$ci_width <- ci_df_19$upper - ci_df_19$lower

votes_per_leg_19 <- rowSums(!is.na(rc_19$votes))
votes_df_19 <- data.frame(
  legis.name = names(votes_per_leg_19),
  n_votes    = as.numeric(votes_per_leg_19),
  stringsAsFactors = FALSE
)
ci_df_19 <- merge(ci_df_19, data19[, c("legis.name", "partito")],
                  by = "legis.name", all.x = TRUE)
ci_df_19 <- merge(ci_df_19, votes_df_19, by = "legis.name", all.x = TRUE)

low_voters_19 <- ci_df_19 %>%
  group_by(partito) %>%
  slice_min(order_by = n_votes, n = 3, with_ties = FALSE) %>%
  ungroup()

set.seed(42)
low_voters_19$y_jitter <- as.numeric(as.factor(low_voters_19$partito)) +
  runif(nrow(low_voters_19), -0.2, 0.2)

ci_scale <- 15

ggplot(low_voters_19) +
  geom_vline(xintercept = 0, linewidth = 1.2, color = "black") +
  geom_point(aes(x = mean, y = y_jitter, size = ci_width * ci_scale),
             color = "grey70", alpha = 0.4) +
  geom_point(aes(x = mean, y = y_jitter, color = partito), size = 2.5) +
  scale_color_manual(values = master_colors) +
  scale_size_identity() +
  scale_y_continuous(
    breaks = seq_along(levels(as.factor(low_voters_19$partito))),
    labels = levels(as.factor(low_voters_19$partito))) +
  theme_minimal() +
  labs(title    = "Posterior Uncertainty — 3 Least-Voting Members per Party (XIX)",
       subtitle = "Grey circle radius = 95% CI half-width | Colored dot = posterior mean",
       x = "← Left          Right →", y = "", color = "Party") +
  guides(size = "none") +
  theme(legend.position = "bottom")
```

To further discuss the relationship between the richness of data and confidence intervals 
below we picture the 95% CI width against the number of votes cast. 
The relationship is consistent with expectations: legislators with
very few recorded votes display substantially wider credible intervals, as their
posteriors remain prior-dominated. Beyond a moderate participation threshold,
however, additional votes contribute diminishing returns to precision — CI width
stabilises and the relationship flattens. The slight upward curvature can be imputed
to loess edge effects. 

```{r plot-h-19, fig.cap="Each point is one legislator. The black LOESS curve shows the average relationship between vote count and posterior precision across the full chamber."}
ci_all_19 <- data.frame(
  legis.name = rownames(id_19$xbar),
  mean       = as.numeric(id_19$xbar[, 1]),
  lower      = apply(id_19$x[, , 1], 2, quantile, 0.025),
  upper      = apply(id_19$x[, , 1], 2, quantile, 0.975),
  stringsAsFactors = FALSE
)
ci_all_19$ci_width <- ci_all_19$upper - ci_all_19$lower
ci_all_19 <- merge(ci_all_19, votes_df_19, by = "legis.name", all.x = TRUE)
ci_all_19 <- merge(ci_all_19, data19[, c("legis.name", "partito")],
                   by = "legis.name", all.x = TRUE)

ggplot(ci_all_19, aes(x = n_votes, y = ci_width)) +
  geom_point(aes(color = partito), alpha = 0.5, size = 1.8) +
  geom_smooth(method = "loess", se = FALSE, 
              color = "black", linewidth = 0.8) +
  scale_color_manual(values = master_colors) +
  theme_minimal() +
  labs(title    = "Posterior Precision vs Data Availability — XIX Legislature",
       subtitle = "More votes → narrower credible interval",
       x = "Number of Votes Cast", y = "95% CI Width", color = "Party") +
  theme(legend.position = "bottom")
```

---

## Results: XVIII Legislature (2018–2022)

The XVIII legislature presents a considerably noisier picture. Three
distinct governments formed over its course: the M5S-Lega coalition
(June 2018 – September 2019), the M5S-PD-IV coalition under Conte II
(September 2019 – February 2021), and the broad technocratic government
under Mario Draghi (February 2021 – October 2022). This instability means
that the left-right signal in the vote record is substantially noisier than
in the XIX legislature, as the same party (most notably M5S) voted with very
different coalitions at different points in time.

```{r load-18, cache=TRUE}
load("session_18.RData")
load("sensitivity_18.RData")

parties_18 <- sort(unique(data18$partito))
party_colors_18 <- setNames(
  scales::viridis_pal(option = "turbo")(length(parties_18)),
  parties_18
)
```

### Ideological Distribution

```{r plot-a-18, fig.cap="Posterior mean ideal points for all XVIII legislature members."}
plot_df_18 <- data.frame(
  legis.name = rownames(id_18$xbar),
  Estimate   = as.numeric(id_18$xbar[, 1]),
  stringsAsFactors = FALSE
)
plot_df_18 <- merge(plot_df_18, data18[, c("legis.name", "partito")],
                    by = "legis.name", all.x = TRUE)

set.seed(42)
ggplot(plot_df_18, aes(x = Estimate, y = 0, color = partito)) +
  geom_jitter(size = 2, alpha = 0.65, height = 0.15) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50") +
  geom_vline(xintercept = 0, linewidth = 0.8, color = "black") +
  scale_color_manual(values = party_colors_18) +
  theme_minimal() +
  labs(title    = "Ideological Distribution — XVIII Legislature",
       subtitle = "Posterior means | Anchors: Fratoianni = -1, Lollobrigida = +1",
       x = "← Left          Right →", y = "", color = "Party") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(), legend.position = "bottom")
```

### Bill Parameters

```{r plot-d-18, fig.width=12, fig.height=5}
beta_df_18 <- data.frame(
  vote_id = 1:nrow(id_18$betabar),
  beta    = id_18$betabar[, 1],
  alpha   = id_18$betabar[, 2]
)
noisy_pct_18 <- round(mean(abs(beta_df_18$beta) < 0.5) * 100, 1)

p_hist_18 <- ggplot(beta_df_18, aes(x = beta)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "steelblue", color = "white", alpha = 0.6) +
  geom_density(color = "black", linewidth = 0.8, adjust = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  theme_minimal() +
  labs(title    = "Distribution of β (XVIII)",
       subtitle = paste0("Votes with |β| < 0.5: ", noisy_pct_18, "%"),
       x = "β (Discrimination)", y = "Density")

hull_idx_18 <- chull(beta_df_18$beta, beta_df_18$alpha)
hull_df_18  <- beta_df_18[hull_idx_18, ]

p_scatter_18 <- ggplot(beta_df_18, aes(x = beta, y = alpha)) +
  geom_polygon(data = hull_df_18, fill = "steelblue", alpha = 0.15,
               color = "steelblue", linewidth = 0.5) +
  geom_point(alpha = 0.3, size = 1.2, color = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.6) +
  theme_minimal() +
  labs(title    = "β vs α (XVIII)",
       subtitle = "Each point is one roll-call vote",
       x = "β (Discrimination)", y = "α (Difficulty)")

p_hist_18 + p_scatter_18
```
This legislature shows a substantially higher share of uninformative votes, with
`r noisy_pct_18`% of bills having $|\beta| < 0.5$. The high share of near-zero discrimination parameters in this legislature may
suggest the need for an additional ideological dimension beyond left-right —
possibly reflecting the cross-cutting coalition dynamics that characterised the
period.

The shift of the M5S from a centrist to a left-leaning position across the two
legislatures can be attributed not only to a genuine ideological realignment, but
also to a statistical artefact of their governing alliance with the Lega: during
that period, M5S voted in favour of bills anchored further to the center-right,
which pulls their estimated ideal point toward the center in the XVIII legislature
and exaggerates the apparent leftward shift when comparing across legislatures.


### Convergence Diagnostics

Convergence is substantially better in this legislature compared to the previous
one, reflecting the larger number of votes available for estimation. Additionally,
the alternation of governing coalitions across legislatures introduces greater
variance in voting behaviour: legislators who shift between majority and
opposition roles over time cast votes from both positions, providing richer
ideological signal and allowing the model to place them more precisely on the
left-right scale. This is further corroborated by the sensitivity to bill prior
specification plot, where — although a slight pull toward the center remains —
the relationship between tight and default prior estimates approaches the 45°
line much more closely, indicating that the data are sufficiently informative to
largely override the prior specification.

```{r geweke-18, cache=TRUE}
leg_18   <- as.mcmc(id_18$x[, , 1])
beta_18  <- as.mcmc(id_18$beta[, , 1])
alpha_18 <- as.mcmc(id_18$beta[, , 2])

gw_leg_18   <- geweke.diag(leg_18)
gw_beta_18  <- geweke.diag(beta_18)
gw_alpha_18 <- geweke.diag(alpha_18)

geweke_df_18 <- data.frame(
  Parameter = c("Ideal Points", "β (Discrimination)", "α (Difficulty)"),
  Total     = c(id_18$n, id_18$m, id_18$m),
  Failed    = c(sum(abs(gw_leg_18$z)   > 1.96),
                sum(abs(gw_beta_18$z)  > 1.96),
                sum(abs(gw_alpha_18$z) > 1.96)),
  Pass_Rate = paste0(round((1 - c(
                mean(abs(gw_leg_18$z)   > 1.96),
                mean(abs(gw_beta_18$z)  > 1.96),
                mean(abs(gw_alpha_18$z) > 1.96))) * 100, 1), "%")
)

knitr::kable(geweke_df_18,
             col.names = c("Parameter", "Total", "Failed (|Z| > 1.96)",
                           "Pass Rate"),
             caption = "Geweke Convergence Diagnostics — XVIII Legislature (1D)")
```



```{r plot-c-18, fig.width=12, fig.height=5}
ideal_comp_18 <- data.frame(
  legis.name = rownames(id_1_18$xbar),
  Tight      = as.numeric(id_1_18$xbar[, 1]),
  Default    = as.numeric(id_2_18$xbar[, 1]),
  Flat       = as.numeric(id_3_18$xbar[, 1]),
  stringsAsFactors = FALSE
)
ideal_comp_18 <- merge(ideal_comp_18, data18[, c("legis.name", "partito")],
                       by = "legis.name", all.x = TRUE)

p_c1_18 <- ggplot(ideal_comp_18, aes(x = Tight, y = Default, color = partito)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = FALSE, color = "red",
              linewidth = 0.8, aes(group = 1)) +
  scale_color_manual(values = party_colors_18, guide = "none") +
  theme_minimal() +
  labs(title = "Tight (Var=1) vs Default (Var=25)",
       x = "Ideal Point (Tight)", y = "Ideal Point (Default)")

p_c2_18 <- ggplot(ideal_comp_18, aes(x = Default, y = Flat, color = partito)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = FALSE, color = "red",
              linewidth = 0.8, aes(group = 1)) +
  scale_color_manual(values = party_colors_18) +
  theme_minimal() +
  labs(title = "Default (Var=25) vs Flat (Var=625)",
       x = "Ideal Point (Default)", y = "Ideal Point (Flat)")

p_c1_18 + p_c2_18 +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  plot_annotation(
    title    = "Sensitivity to Bill Prior Specification — XVIII Legislature",
    subtitle = "Dashed line: identity | Red line: OLS fit")
```

### Data, Uncertainty and Precision
As expected, legislators with few recorded votes have their posteriors dominated
by the prior, with ideal point estimates clustered near the center of the scale
and very wide, overlapping credible intervals. For most of these MPs the model
cannot meaningfully distinguish between individuals — a reminder that the
estimates are informative at the party level, but should be interpreted
cautiously at the individual level for low-participation legislators.

The overall trend shown below is consistent with expectations: there is a clear
negative relationship between the number of votes cast and the width of the
credible interval, which flattens at around 100 votes — suggesting that beyond
this threshold, additional votes contribute little further precision to the
estimates.

```{r plot-b-18, fig.height=8}
ci_df_18 <- data.frame(
  legis.name = rownames(id_18$xbar),
  mean       = as.numeric(id_18$xbar[, 1]),
  lower      = apply(id_18$x[, , 1], 2, quantile, 0.025),
  upper      = apply(id_18$x[, , 1], 2, quantile, 0.975),
  stringsAsFactors = FALSE
)
ci_df_18$ci_width <- ci_df_18$upper - ci_df_18$lower

votes_per_leg_18 <- rowSums(!is.na(rc_18$votes))
votes_df_18 <- data.frame(
  legis.name = names(votes_per_leg_18),
  n_votes    = as.numeric(votes_per_leg_18),
  stringsAsFactors = FALSE
)
ci_df_18 <- merge(ci_df_18, data18[, c("legis.name", "partito")],
                  by = "legis.name", all.x = TRUE)
ci_df_18 <- merge(ci_df_18, votes_df_18, by = "legis.name", all.x = TRUE)

low_voters_18 <- ci_df_18 %>%
  group_by(partito) %>%
  slice_min(order_by = n_votes, n = 3, with_ties = FALSE) %>%
  ungroup()

set.seed(42)
low_voters_18$y_jitter <- as.numeric(as.factor(low_voters_18$partito)) +
  runif(nrow(low_voters_18), -0.2, 0.2)

ci_scale <- 15

ggplot(low_voters_18) +
  geom_vline(xintercept = 0, linewidth = 1.2, color = "black") +
  geom_point(aes(x = mean, y = y_jitter, size = ci_width * ci_scale),
             color = "grey70", alpha = 0.4) +
  geom_point(aes(x = mean, y = y_jitter, color = partito), size = 2.5) +
  scale_color_manual(values = party_colors_18) +
  scale_size_identity() +
  scale_y_continuous(
    breaks = seq_along(levels(as.factor(low_voters_18$partito))),
    labels = levels(as.factor(low_voters_18$partito))) +
  theme_minimal() +
  labs(title    = "Posterior Uncertainty — 3 Least-Voting Members per Party (XVIII)",
       subtitle = "Grey circle radius = 95% CI half-width | Colored dot = posterior mean",
       x = "← Left          Right →", y = "", color = "Party") +
  guides(size = "none") +
  theme(legend.position = "bottom")
```


```{r plot-h-18}
ci_all_18 <- data.frame(
  legis.name = rownames(id_18$xbar),
  mean       = as.numeric(id_18$xbar[, 1]),
  lower      = apply(id_18$x[, , 1], 2, quantile, 0.025),
  upper      = apply(id_18$x[, , 1], 2, quantile, 0.975),
  stringsAsFactors = FALSE
)
ci_all_18$ci_width <- ci_all_18$upper - ci_all_18$lower
ci_all_18 <- merge(ci_all_18, votes_df_18, by = "legis.name", all.x = TRUE)
ci_all_18 <- merge(ci_all_18, data18[, c("legis.name", "partito")],
                   by = "legis.name", all.x = TRUE)

ggplot(ci_all_18, aes(x = n_votes, y = ci_width, color = partito)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(method = "loess", se = FALSE, color = "black",
              linewidth = 0.8, aes(group = 1)) +
  scale_color_manual(values = party_colors_18) +
  theme_minimal() +
  labs(title    = "Posterior Precision vs Data Availability — XVIII Legislature",
       subtitle = "More votes → narrower credible interval",
       x = "Number of Votes Cast", y = "95% CI Width", color = "Party") +
  theme(legend.position = "bottom")
```

---

## Cross-Legislature Analysis: Party Switchers

Legislators present in both the XVIII and XIX legislatures who changed party
affiliation offer a unique opportunity to assess whether the model captures
individual ideology or merely party-line voting. If ideal points are stable
across legislatures despite party changes, it suggests the estimates reflect
genuine individual preferences. If they shift substantially, it may indicate
that voting behaviour — and therefore estimated ideal points — is driven more
by coalition dynamics than individual ideology.

Since both legislatures use the same anchors (Fratoianni = $-1$,
Lollobrigida = $+1$), the two scales are directly comparable.

```{r plot-g, fig.height=20, fig.cap="Grey dot = XVIII position. Arrow tip = XIX position. Legislators sorted by magnitude of shift."}
ip_18 <- data.frame(
  legis.name = rownames(id_18$xbar),
  ip_18      = as.numeric(id_18$xbar[, 1]),
  stringsAsFactors = FALSE
)

ip_19 <- data.frame(
  legis.name = rownames(id_19$xbar),
  ip_19      = as.numeric(id_19$xbar[, 1]),
  stringsAsFactors = FALSE
)

party_18 <- data18[, c("legis.name", "partito")] %>%
  rename(partito_18 = partito) %>% distinct()
party_19 <- data19[, c("legis.name", "partito")] %>%
  rename(partito_19 = partito) %>% distinct()

switchers <- party_18 %>%
  inner_join(party_19, by = "legis.name") %>%
  filter(partito_18 != partito_19) %>%
  inner_join(ip_18, by = "legis.name") %>%
  inner_join(ip_19, by = "legis.name") %>%
  mutate(delta     = ip_19 - ip_18,
         direction = ifelse(delta > 0, "Moved Right", "Moved Left"),
         abs_delta = abs(delta)) %>%
  arrange(desc(abs_delta))

switchers$label <- paste0(switchers$legis.name, "\n(",
                           switchers$partito_18, " → ",
                           switchers$partito_19, ")")
ci_18 <- data.frame(
  legis.name = rownames(id_18$xbar),
  lower_18   = apply(id_18$x[, , 1], 2, quantile, 0.025),
  upper_18   = apply(id_18$x[, , 1], 2, quantile, 0.975),
  stringsAsFactors = FALSE
)

switchers <- switchers %>%
  inner_join(ci_18, by = "legis.name") %>%
  # Primary: XIX position must fall outside XVIII 95% CI
  filter(ip_19 < lower_18 | ip_19 > upper_18) 
  
ggplot(switchers) +
  geom_segment(aes(x = ip_18, xend = ip_19,
                   y = reorder(label, delta), yend = reorder(label, delta),
                   color = direction),
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
               linewidth = 0.9) +
  geom_point(aes(x = ip_18, y = reorder(label, delta)),
             shape = 16, size = 2.5, color = "gray40") +
  geom_vline(xintercept = 0, linewidth = 1, color = "black") +
  scale_color_manual(values = c("Moved Right" = "#E41A1C",
                                "Moved Left"  = "#377EB8")) +
  theme_minimal() +
  labs(title    = "Ideal Point Shift for Party Switchers (XVIII → XIX)",
       subtitle = "Grey dot = XVIII position | Arrow tip = XIX position",
       x = "← Left          Right →", y = "", color = "Direction") +
  theme(legend.position    = "bottom",
        axis.text.y        = element_text(size = 8),
        panel.grid.major.y = element_line(color = "gray90"))
```



---

## Conclusions {#Conclusions}
Due to strong party whipping, the estimation of ideal points for Italian
legislators is considerably more challenging than in the US context, where
members of Congress have historically enjoyed greater freedom in their voting
behaviour — though increasing polarisation has reduced this in recent years.
Nevertheless, the model remains a valid tool for extracting aggregate party
positions and inter-party distances, even when individual-level estimates are
imprecise.

The analysis above also highlights how data richness is critical for the
reliable performance of simulation-based statistical tools: convergence,
precision, and robustness to prior specification all improve substantially as
the number of observed votes increases — even in the face of what might appear
as noise, as the XVIII legislature illustrates.